The interactions of rational, pragmatic agents\\ lead to efficient language structure and use

### 2019-02-01 paper/paper.tex
1st round FD updates

Zipf articulated what is now considered a \emph{functionalist} approach
to language science -- analyzing language structure and use in terms of
efficiency. Such an approach might reframe our opening question as
follows: how does having property \textit{x} make using language
\(\ell\) more or less useful for communication? Among others, this
efficiency-based framing has produced a rich set of theoretical and
empirical targets exploring semantic typology (Regier, Kemp, \& Kay,
2015), properties such as ambiguity (Piantadosi, Tily, \& Gibson, 2011)
and compositionality (Kirby, Griffiths, \& Smith, 2014), and the
efficient use of reduction and redundancy in production (Genzel \&
Charniak, 2002; Levy \& Jaeger, 2007). These projects typically employ
information-theoretic principles to characterize efficiency in natural
or simulated language corpora. We adopt a different approach in this
work, deriving a measure of efficiency from first principles of
\textit{efficient language use} and show how this can lead to
communicatively efficient systems.\par

@@

Adopting reference games as a basic unit of analysis suggests that
optimization may take place at the level of conversation. Importantly,
Zipf's conception of speaker and listener effort may be connected to how
language is used; in particular, whether interlocutors engage in
pragmatic reasoning during conversation. Under a Gricean treatment of
pragmatics, speakers and listeners follow a set of conversational maxims
in which they cooperate to transfer information (Grice, 1975). These
maxims appear to emerge from efficiency concerns, however (Horn, 1984).
We formalize this connection -- showing how system-level efficiencies
can emerge from local interaction behavior of pragmatic agents.
Intuitively, our claim is that to understand an ``efficient'' property
of a system it is essential that we consider how that property is
\emph{used} efficiently.\par

@@

The Rational Speech Act framework (RSA) is a computational-level theory
of pragmatic language use, which has produced good fit to human
communication behavior across a range of language phenomena (Frank \&
Goodman, 2012; Goodman \& Frank, 2016). RSA is a formalization of
essential Gricean pragmatic principles -- agents reason about one
another and their shared context (Grice, 1975). We adopt RSA as our
representational framework to model Gricean (rational and pragmatic)
speaker-listeners in the reference game setting (see SI and Goodman \&
Frank (2016) for review).\par

@@

Importantly, we assume that events \(e = <u, m, c>\) are sampled
according the to following generative model -- some context occurs in
the world with probability \(P(C=c)\). Within this context, an object
\(m\) occurs with probability \(p(m|c)\). The speaker attempts to refer
to that object by sampling from her conditional distribution
\(S(u|m, c; \ell)\) (i.e. \(e \sim p(c)p(m|c)S(u|m, c; \ell)\)). From
the ingredients described above, it is possible to derive the following
objective between the speaker and listener joint distributions (see SI
2.2 for a complete derivation).

@@

From an information-theoretic perspective this objective is intuitive:
\(H_{cross}\) denotes the Cross-Entropy (CE), a measure of dissimilarity
between two distributions -- the average number of bits required to
communicate under one distribution, given that the ``true'' distribution
differs. In our case, we have an expectation over this term -- the
expected difference between the distributions assumed by the speaker
\(P_{speaker}\) and listener \(P_{listener}\) given a set of contexts
\(C\)\footnote{Note that in the single context case $|C|=1$ this objective is simply the speaker-listener Cross-Entropy over utterance-meaning pairs $H_{cross}[P_{speaker}(u, m; \ell), P_{listener}(u, m; \ell)]$}.
In other words, an ``efficient'' language \(\ell\) minimizes the
distance between what speakers and listeners think.

@@

The task of understanding language is marked by a frequent need to
handle various forms of ambiguity: lexical, syntactic, among others
(Wasow, Perfors, \& Beaver, 2005). The ubiquity of this property,
however, has been argued to provide evidence that languages have not
been optimized for
communication\footnote{Chomsky's (2002) claim that "... language is not well designed, because you have such properties as ambiguity" is analogous to the Zipfian optimal Listener language $\ell_{listener}^*$.}
(Chomsky, 2002).

@@

In the following experiments we explore two aspects of Piantadosi's et
al.'s claim. In Simulation 1, we examine the efficient language
\emph{structure} aspect of their claim, exploring when the optimal
linguistic system \(\ell^*\) is most likely to contain ambiguous lexical
items. In Simulation 2, we explore an efficient language \emph{use}
aspect of the claim -- at what point in a conversation is it useful for
a speaker to use ambiguous lexical material?\par

@@

\caption[Panel (A) Vertical axis shows the proportion of optimal languages containing ambiguity]{Panel (A) Vertical axis shows the proportion of optimal languages containing ambiguity. Horizontal axis shows the number of context-sizes (1-4). Optimal language under CE objective (red). Speaker-optimal (blue). Listener-optimal (green). Error bars represent 95 percent confidence intervals. Panel (B), example CE-optimal language (ambiguous) under in a four-condition context. Panel (C), example CE-optimal language in a single-condition context.}\label{fig:plot-optimal-langs}
\end{figure*}
\end{CodeChunk}

@@

We show that ambiguity is an efficient property under our CE objective
in the reference game setting. We proceed by generating languages with
different amounts of contextual support (varying the size of \(|C|\)).
We search the space of languages, examining whether ones which minimize
our objective contain ambiguity. If context leads to more efficient
communication, then optimal languages under our objective should be more
likely to be ambiguous as the amount of context increases.\par

@@

For comparison, we also examine properties of optimal languages under
two additional objectives. Zipf (1949) proposed that the speaker-optimal
language \(\ell_{speaker}^*\) would minimize speaker effort and the
listener-optimal language \(\ell_{listener}^*\) would minimize listener
effort. We define these objectives using the first and second half of
equation 1 (see SI section 2.2. for more detail).\par

@@

In Simulation 1 we explored the degree to which ambiguity is an
efficient property of languages when communication is contextualized.
Figure 1 plots the proportion of optimal languages under each objective
as a function of number of contexts. The red line shows that as the
number of contexts increases, so does the probability that an optimal
language \(\ell^*_{cross}\) contains ambiguity (at least one utterance
maps to two meanings) under our CE objective. For comparison we also
plot the proportion of speaker-optimal \(\ell^*_{speaker}\) (blue line)
and listener-optimal \(\ell^*_{listener}\) (green line) languages that
contain ambiguity. In line with Zipf's predictions, if languages are
designed only to minimize speaker effort then optimal languages always
contain ambiguity. If languages are designed to minimize listener effort
then ambiguity is always avoided.\par

@@

We consider consider a single language \(\ell\), which contains both
ambiguous and unambiguous utterances. We assume ambiguous utterances are
lower cost. Crucially, we do not assume that the listener knows the
particular topic (\(c_{current}\)) of the conversation
\textit{a priori}. Rather, that the listener has knowledge of the set of
possible topics \(C = \{c_1, \dots, c_k\}\), but does not know which one
is currently being used by the speaker. Formally, this means the
listener does not have access to the correct conditional distribution
over meanings \(P(M|C=c_{current})\) at the start of the discourse.\par

Over the course of a discourse \(D\), the listener tries to infer both
the current topic, \(c_{current}\), as well as the particular meaning
\(m\) of a given utterance \(u\). That is, we consider listener and
speaker agents who can track the history of previous utterances \(D\).
Importantly, an agent can attempt to infer the current topic of
conversation \(c_{current}\) using the discourse history \(D\).\par

@@

We vary the degree to which agents can use context for disambiguation.
We consider three types of speaker models. Our \textit{Full pragmatics}
agent, models a speaker who reasons about her listener and also has
complete recall of the set of utterances in the discourse \(D\). This
speaker believes that the listener may not know the current topic
\(c_{current}\) at the start of the discourse, but can infer it over the
discourse. We compare two baseline models. The first, a
\textit{Partial pragmatics} baseline describes a speaker who reasons
about a listener, but assumes they have no access to the discourse
history. The second, a \textit{No pragmatics} speaker does not consider
a listener at all, but produces utterances according to the underlying
langauge semantics (\(\ell\)) and topic probabilities
(\(p(M|C=c_{current}\)) (see SI section 3).\par

@@

Figure 2 (A) shows the empirical probability that a speaker uses an
ambiguous utterance as a function of discourse position. The
\textit{No pragmatic} baseline uses ambiguous utterances frequently and
at a constant rate over the discourse and the \textit{Partial pragmatic}
baseline avoids ambiguous utterances entirely. But, the
\textit{Full pragmatic} model avoids ambiguous material only at the
start of the discourse, employing it increasingly as the discourse
proceeds. Panel (C) tracks our CE objective for each model over the
discourse. Note that the objective decreases for all three models,
primarily driven by the listener updating his belief about the actual
topic (\(P(C=c_{current}|D)\)). However, the objective declines more
quickly under the \textit{Full} and \textit{Partial} pragmatic speakers
as listener agents are better able to infer the correct context.
Additionally, the difference in CE between the \textit{Full}- and
\textit{Partial pragmatic} models at the end of the discourse is driven
by the reduction in speaker costs (see panel B). While speaker effort
remains constant in both \textit{No pragmatic} and
\textit{Partial pragmatic} baselines, effort declines in the
\textit{Full pragmatic} model as she increasingly relies on ambiguous
material later in the discourse.\par

@@

How do the competing pressures of speakers and listeners give rise to
the distributional regularities found in natural language? Zipf (1949)
proposed that the asymmetry between speaker and listener costs gives
rise to a range of properties at the level of the lexicon. We explored
the interactions of rational pragmatic agents as a framework for
understanding efficient language structure and use. We focused on an
argument on the communicative function of ambiguity (Piantadosi et al.,
2011), deriving a novel speaker-listener Cross-Entropy objective for
measuring the efficiency of linguistic systems from first principles of
efficient language use. In a series of simulations we showed that
optimal languages are more likely to contain ambiguous material when
context is informative and that rational pragmatic agents will use
ambiguous material efficiently.\par

A limitation of the current work is an analysis of exactly how this
objective compares to existing measures. For example, previous work has
described competing speaker-listener pressures in terms of a trade-off
of simplicity and informativeness (Kemp \& Regier, 2012) or expressivity
and compressibility (Smith, Tamariz, \& Kirby, 2013). Future work should
assess the degree to which we can derive the same properties as previous
studies using our current framework. In line with this idea, a primary
focus of the current work is to provide a case-study on the productivity
of assessing functionalist theories in repeated reference games between
rational pragmatic agents. We hope this framework can serve as a domain
general tool to assess the range of functionalist theories examining
efficient language-structure and use.\par

>>>>

Zipf articulated what is now considered a \emph{functionalist} approach
to language science -- analyzing language structure and use in terms of
efficiency. Such an approach might reframe our opening question as
follows: how does having property \textit{x} make using language
\(\ell\) more or less useful for communication? Among others, this
efficiency-based framing has produced a rich set of theoretical and
empirical targets exploring semantic typology (Regier, Kemp, \& Kay,
2015), properties such as ambiguity (Piantadosi, Tily, \& Gibson, 2011)
and compositionality (Kirby, Griffiths, \& Smith, 2014), and the
efficient use of reduction and redundancy in production (Genzel \&
Charniak, 2002; Levy \& Jaeger, 2007). These projects typically employ
information-theoretic principles to characterize efficiency in natural
or simulated language corpora.\par

We adopt a different approach in this work in two ways. First we derive
a measure of efficiency from first principles of
\textit{efficient language use} and show how this can lead to
communicatively efficient systems. Second we demonstrate how assumptions
about language-users impacts whether language properties are used
efficiently. In this way, we bridge questions of design and use in a
single framework.\par

@@

Adopting reference games as a basic unit of analysis suggests that
optimization may take place at the level of conversation. Importantly,
Zipf's conception of speaker and listener effort may be connected to how
language is used; in particular, whether interlocutors engage in
pragmatic reasoning during conversation. Under a Gricean treatment of
pragmatics, speakers and listeners follow a set of conversational maxims
in which they cooperate to transfer information (Grice, 1975). These
maxims appear to emerge from efficiency concerns, however (Horn, 1984).
We formalize this connection -- showing how system-level efficiencies
can emerge from local interaction behavior of pragmatic agents. Our
claim is that to understand an ``efficient'' property of a system it is
essential that we consider how that property is \emph{used}
efficiently.\par

@@

The Rational Speech Act framework (RSA) is a computational-level theory
of pragmatic language use, which has produced good fit to human
communication behavior across a range of language phenomena (Frank \&
Goodman, 2012; Goodman \& Frank, 2016). RSA is a formalization of
essential Gricean pragmatic principles -- agents reason about one
another and their shared context (Grice, 1975). We adopt RSA as our
representational framework to model Gricean (rational and pragmatic)
speaker-listeners in the reference game setting (see
\href{https://github.com/benpeloquin7/zipf_principles/blob/master/paper/supplementary_materials.pdf}{SI}
and Goodman \& Frank (2016) for a review).\par

@@

Importantly, we assume that events \(e = <u, m, c>\) are sampled
according the to following generative model -- some context occurs in
the world with probability \(P(C=c)\). Within this context, an object
\(m\) occurs with probability \(p(m|c)\). The speaker attempts to refer
to that object by sampling from her conditional distribution
\(S(u|m, c; \ell)\) (i.e. \(e \sim p(c)p(m|c)S(u|m, c; \ell)\)). From
the ingredients described above, it is possible to derive the following
objective between the speaker and listener distributions (see
\href{https://github.com/benpeloquin7/zipf_principles/blob/master/paper/supplementary_materials.pdf}{SI 2.1}
for a complete derivation).

@@

From an information-theoretic perspective this objective is intuitive:
\(H_{cross}\) denotes the Cross-Entropy (CE), a measure of dissimilarity
between two distributions -- the average number of bits required to
communicate under one distribution, given that the ``true'' distribution
differs. In our case, we have an expectation over this term -- the
expected difference between the distributions assumed by the speaker
\(P_{speaker}\) and listener \(P_{listener}\) given a set of contexts
\(C\)\footnote{Note that in the single context case $|C|=1$ this objective is simply the speaker-listener Cross-Entropy.}.
In other words, an ``efficient'' language \(\ell\) minimizes the
distance between what speakers and listeners think.

@@

The task of understanding language is marked by a frequent need to
handle various forms of ambiguity: lexical, syntactic, among others
(Wasow, Perfors, \& Beaver, 2005). The ubiquity of this property,
however, has been argued to provide evidence that languages have not
been optimized for communication (Chomsky, 2002).

@@

In the following experiments we explore two aspects of Piantadosi's et
al.'s claim. In Simulation 1, we examine the efficient language
\emph{structure} aspect of their claim, exploring when the optimal
linguistic system \(\ell^*\) is most likely to contain ambiguous
expressions. In Simulation 2, we explore an efficient language
\emph{use} aspect of the claim -- at what point in a conversation is it
useful for a speaker to use ambiguous expressions?\par

@@

\caption[Panel (A) Vertical axis shows the proportion of optimal languages containing ambiguity]{Panel (A) Vertical axis shows the proportion of optimal languages containing ambiguity. Horizontal axis shows the number of context-sizes (1-4). Optimal language under CE objective (red). Speaker-optimal (blue). Listener-optimal (green). Error bars represent 95 percent confidence intervals. Panel (B), example CE-optimal language (ambiguous) under in a four-condition context. Panel (C), example CE-optimal language (unambiguous) in a single-condition context.}\label{fig:plot-optimal-langs}
\end{figure*}
\end{CodeChunk}

@@

We show that ambiguity is an efficient property under our CE objective
in the reference game setting. We proceed by generating languages with
different amounts of contextual support (varying the size of \(|C|\)).
We search the space of languages, examining whether ones which minimize
our objective contain ambiguity. If context leads to more efficient
communication, then optimal languages should be more likely to be
ambiguous as the amount of context increases.\par

@@

For comparison, we also examine properties of optimal languages under
two additional objectives. Zipf (1949) proposed that the speaker-optimal
language \(\ell_{speaker}^*\) would minimize speaker effort and the
listener-optimal language \(\ell_{listener}^*\) would minimize listener
effort. We define these objectives using the first and second half of
equation 1 (see
\href{https://github.com/benpeloquin7/zipf_principles/blob/master/paper/supplementary_materials.pdf}{SI 2.2.}).\par

@@

In Simulation 1 we explored the degree to which ambiguity is an
efficient property of languages when communication is contextualized.
Figure 2, panel (A) plots the proportion of optimal languages under each
objective as a function of number of contexts. The red line shows that
as the number of contexts increases, so does the probability that an
optimal language \(\ell^*_{cross}\) contains ambiguity (at least one
utterance maps to two meanings) under our CE objective. For comparison
we also plot the proportion of speaker-optimal \(\ell^*_{speaker}\)
(blue line) and listener-optimal \(\ell^*_{listener}\) (green line)
languages that contain ambiguity. In line with Zipf's predictions, if
languages are designed only to minimize speaker effort then optimal
languages always contain ambiguity. If languages are designed to
minimize listener effort then ambiguity is always avoided.\par

@@

We consider a single language \(\ell\), which contains both ambiguous
and unambiguous utterances. We assume ambiguous utterances are lower
cost. Crucially, we do not assume that the listener knows the particular
topic (\(c_{current}\)) of the conversation \textit{a priori}. Rather,
that the listener has knowledge of the set of possible topics
\(C = \{c_1, \dots, c_k\}\), but does not know which one is currently
being used by the speaker. Formally, this means the listener does not
have access to the correct conditional distribution over meanings
\(P(M|C=c_{current})\) at the start of the discourse.\par

Over the course of a discourse \(D\), the listener tries to infer both
the current topic, \(c_{current}\), as well as the particular meaning
\(m\) of a given utterance \(u\). That is, we consider agents who can
track the history of previous utterances \(D\). Importantly, an agent
can attempt to infer the current topic of conversation \(c_{current}\)
using the discourse history \(D\).\par

@@

We vary the degree to which agents can use context for disambiguation.
We consider three types of speaker models. Our \textit{Full pragmatics}
agent, models a speaker who reasons about her listener and also has
complete recall of the set of utterances in the discourse \(D\). This
speaker believes that the listener may not know the current topic
\(c_{current}\) at the start of the discourse, but can infer it over the
discourse. We compare two baseline models. The first, a
\textit{Partial pragmatics} baseline describes a speaker who reasons
about a listener, but assumes they have no access to the discourse
history. The second, a \textit{No pragmatics} baseline speaker does not
consider a listener at all, but produces utterances according to the
underlying langauge semantics (\(\ell\)) and topic probabilities
(\(p(M|C=c_{current}\)) (see
\href{https://github.com/benpeloquin7/zipf_principles/blob/master/paper/supplementary_materials.pdf}{SI 3}).\par

@@

Figure 2 (A) shows the empirical probability that a speaker uses an
ambiguous utterance as a function of discourse position. The
\textit{No pragmatic} baseline uses ambiguous utterances frequently and
at a constant rate over the discourse. The \textit{Partial pragmatic}
baseline avoids ambiguous utterances entirely. But, the
\textit{Full pragmatic} model avoids ambiguous material only at the
start of the discourse, employing it increasingly as the discourse
proceeds. Panel (C) tracks our CE objective for each model over the
discourse. Note that the objective decreases for all three models,
primarily driven by the listener updating his belief about the actual
topic (\(P(C=c_{current}|D)\)). However, the objective declines more
quickly under the \textit{Full} and \textit{Partial} pragmatic speakers
as listener agents are better able to infer the correct context.
Additionally, the difference in CE between the \textit{Full}- and
\textit{Partial pragmatic} models at the end of the discourse is driven
by the reduction in speaker costs (see panel B). While speaker effort
remains constant in both \textit{No pragmatic} and
\textit{Partial pragmatic} baselines, effort declines in the
\textit{Full pragmatic} model as she increasingly relies on ambiguous
material later in the discourse.\par

@@

How do the competing pressures of speakers and listeners give rise to
the distributional regularities found in natural language? Zipf (1949)
proposed that the asymmetry between speaker and listener costs gives
rise to a range of properties at the level of the lexicon. We explored
the interactions of rational pragmatic agents as a framework for
understanding efficient language structure and use. We focused on an
argument on the communicative function of ambiguity (Piantadosi et al.,
2011), deriving a novel speaker-listener Cross-Entropy objective for
measuring the efficiency of linguistic systems from first principles of
efficient language use. In Simulation 1 we showed that optimal languages
are more likely to contain ambiguous material when context is
informative. In Simulation 2 we showed how rational pragmatic agents use
ambiguous material efficiently in conversation.\par

A limitation of the current work is an analysis of exactly how the CE
objective compares to existing measures. For example, previous work has
described competing speaker-listener pressures in terms of a trade-off
of simplicity and informativeness (Kemp \& Regier, 2012) or expressivity
and compressibility (Smith, Tamariz, \& Kirby, 2013) to explain
linguistic regularities. Future work should assess the degree to which
we can derive the same properties as previous studies using our current
framework. In line with this idea, a primary focus of the current work
is to provide a case-study on the productivity of assessing
functionalist theories in repeated reference games between rational
pragmatic agents. We hope this framework can serve as a domain general
tool to assess the range of functionalist theories examining efficient
language-structure and use.\par

\vspace{1em}

\fbox{\parbox[b][][c]{7.3cm}{\centering {Code and SI are available at this \href{https://github.com/benpeloquin7/zipf_principles/}{link}.}}}
\vspace{1em} \noindent
